\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-11-19 paper/main.tex

To obtain a clean preliminary estimate of the consequences on memory of
randomly varying the font colors and locations of presented words (versus
holding the font color fixed at black, and holding the display locations fixed
at the center of the display) we compared participants' performance on the
\textit{feature-rich} and \textit{reduced} experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In the feature-rich
condition the words' colors and locations varied randomly across words, and in
the reduced condition words were always presented in black, at the center of
the display. Aggregating across all lists for each participant, we found no
difference in recall accuracy (i.e., the proportions of correctly recalled
words) for feature-rich versus reduced lists ($t(126) = -0.290, p = 0.772,
\mathrm{Cohen's}~d~(d) = -0.051,
\mathrm{bootstrap~estimated~95\%~confidence~interval}~(\mathrm{CI}) = [-2.387,
1.768]$). However, participants in the feature-rich condition clustered their
recalls substantially more along every dimension we examined (temporal
clustering: $t(126) = 10.632, p < 0.001, d = 1.882, \mathrm{CI} = [7.786,
14.386]$; semantic category clustering: $t(126) = 10.148, p < 0.001, d = 1.796,
\mathrm{CI} = [7.324, 13.778] $; size clustering: $t(126) = 12.033, p < 0.001,
d = 2.129, \mathrm{CI} = [9.030, 15.918]$; word length clustering: $t(126) =
10.720, p < 0.001, d = 1.897, \mathrm{CI} = [7.442, 15.174]$; first letter
clustering: $t(126) = 6.679, p < 0.001, d = 1.182, \mathrm{CI} = [4.490,
9.611]$; see \textit{Permutation-corrected feature clustering scores} for more
information about how we quantified each participant's clustering tendencies.)
Taken together, these comparisons suggest that adding new features changes how
participants organize their memories of studied words, even when those new
features are independent of the words themselves and even when the new features
vary randomly across words. We found no evidence that those additional
uninformative features were distracting (in terms of their impact on memory
performance), but they did affect participants' recall dynamics (measured via
their clustering scores).

A core assumption of our approach is that each participant organizes their
memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order, in their recall sequences (and,
presumably, organize in memory), the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints
computed for different participants. We reasoned that the feature-rich
condition would provide the best opportunity to test this assumption, since the
clustering scores would not be potentially confounded by order manipulations.
To test our ``unique memory fingerprint'' assumption, we compared the
similarity (correlation) between the fingerprint from a single list (from one
participant) and (a) the average fingerprint from all other lists from the same
participant versus (b) the average fingerprint from each other participant
(across all of their lists). Repeating this procedure for all lists and
participants, we found that participants' fingerprints on a held-out list are
reliably more similar to the same participant's fingerprints on other lists
than to other participants' fingerprints ($t(70280) = 5.077, p < 0.001, d =
0.162, \mathrm{CI} = [3.086, 6.895]$). This suggests that participants'
fingerprints are stable across lists, and that each participant's fingerprint
is unique to them.

We next asked whether adding these incidental visual features to later lists
(after the participants had already studied impoverished lists), or removing
the visual features from later lists (after the participants had already
studied visually diverse lists) might affect memory performance. In other
words, we sought to test for potential effects of changing the ``richness'' of
participants' experiences over time. All participants studied and recalled a
total of 16 lists; we defined \textit{early} lists as the first eight lists and
\textit{late} lists as the last eight lists each participant encountered. To
help interpret our results, we compared participants' memories on early versus
late lists in the above feature-rich and reduced conditions. Participants in
both conditions remembered more words on early versus late lists (feature-rich:
$t(66) = 4.553, p < 0.001, d = 0.233, \mathrm{CI} = [2.427, 7.262]$; reduced:
$t(60) = 2.434, p = 0.018, d = 0.134, \mathrm{CI} = [0.493, 4.910]$).
Participants in the feature-rich (but not reduced) conditions exhibited more
temporal clustering on early versus late lists (feature-rich: $t(66) = 2.268, p
= 0.027, d = 0.181, \mathrm{CI} = [0.437, 4.425]$; reduced: $t(60) = 0.986, p =
0.328, d = 0.061, \mathrm{CI} = [-0.897, 3.348]$). And participants in both
conditions tended to exhibit more semantic clustering on early versus late
lists (feature-rich, category: $t(66) = 3.684, p < 0.001, d = 0.220,
\mathrm{CI} = [1.733, 5.732]$; feature-rich, size: $t(66) = 1.629, p = 0.108, d
= 0.100, \mathrm{CI} = [-0.207, 3.905] $; reduced, category: $t(60) = 2.755, p
= 0.008, d = 0.177, \mathrm{CI} = [0.761, 5.189]$; reduced, size: $t(60) =
3.081, p = 0.003, d = 0.201, \mathrm{CI} = [1.210, 5.326]$). Participants in
the reduced (but not feature-rich) conditions tended to exhibit more
lexicographic clustering on early versus late lists (feature-rich, word length:
$t(66) = -0.100, p = 0.921, d = -0.010, \mathrm{CI} = [-2.217, 1.899]$; feature
rich, first letter: $t(66) = 0.412, p = 0.681, d = 0.045, \mathrm{CI} =
[-1.645, 2.461]$; reduced, word length: $t(60) = 3.762, p < 0.001, d = 0.261,
\mathrm{CI} = [1.604, 6.821]$; reduced, first letter: $t(60) = 1.721, p =
0.090, d = 0.175, \mathrm{CI} = [-0.138, 4.098]$). Taken together, these
comparisons suggest that even when the presence or absence of incidental visual
features is stable across lists, participants still exhibit some differences in
their performance and memory organization tendencies for early versus late
lists.

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to exhibit stronger clustering effects on feature-rich
(versus reduced) lists, and (b) participants tended to remember more words and
exhibit stronger clustering effects on early (versus late) lists, we expected
these early versus late differences to be enhanced in the reduced (early)
condition and diminished in the reduced (late) condition. However, to our
surprise, participants in \textit{neither} condition exhibited reliable early
versus late differences in accuracy (reduced (early): $t(41) = 1.499, p =
0.141, d = 0.098, \mathrm{CI} = [-0.345, 3.579]$; reduced (late): $t(40) =
1.462, p = 0.152, d = 0.121, \mathrm{CI} = [-0.376, 2.993]$), temporal
clustering (reduced (early): $t(41) = 0.857, p = 0.396, d = 0.068, \mathrm{CI}
= [-1.012, 2.896]$; reduced (late): $t(40) = 1.244, p = 0.221, d = 0.128,
\mathrm{CI} = [-0.894, 3.088]$), nor feature-based clustering (reduced (early),
category: $t(41) = 0.707, p = 0.484, d = 0.068, \mathrm{CI} = [-1.314, 2.830]$;
reduced (early), size: $t(41) = 0.803, p = 0.427, d = 0.079, \mathrm{CI} =
[-1.142, 2.953]$; reduced (early), length: $t(41) = 0.461, p = 0.648, d =
0.060, \mathrm{CI} = [-1.545, 2.462]$; reduced (early), first letter: $t(41) =
0.781, p = 0.439, d = 0.101, \mathrm{CI} = [-1.039, 2.881]$; reduced (late),
category: $t(40) = 0.101, p = 0.920, d = 0.009, \mathrm{CI} = [-1.776, 2.307]$;
reduced (late), size: $t(40) = 0.555, p = 0.582, d = 0.058, \mathrm{CI} =
[-1.444, 2.274] $; reduced (late), length: $t(40) = 1.482, p = 0.146, d =
0.126, \mathrm{CI} = [-0.444, 3.743]$; reduced (late), first letter: $t(40) =
-0.143, p = 0.887, d = -0.017, \mathrm{CI} = [-2.204, 1.830] $). We
hypothesized that adding or removing the variability in the visual features was
acting as a sort of ``event boundary'' between early and late
lists~\citep[e.g., ][]{ClewEtal19,RadvZack17,RadvCope06}. In prior work, we
(and others) have found that memories formed just after event boundaries can be
enhanced~\citep[e.g., due to less contextual interference between pre- and
post-boundary items;][]{MannEtal16, PettEtal16, GoldEtal17, FlorEtal17}.

We found that \textit{adding} incidental visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature-rich versus reduced (early):
$t(107) = -2.230, p = 0.028, d = -0.439, \mathrm{CI} = [-4.252, -0.229]$;
reduced versus reduced (early): $t(101) = -2.045, p = 0.043, d = -0.410,
\mathrm{CI} = [-3.826, 0.112]$; also see Fig.~\accuracyByList A). However,
\textit{subtracting} irrelevant visual features on later lists that
\textit{had} been present on early lists (as in the reduced (late) condition)
did not appear to impact recall performance (accuracy for feature-rich versus
reduced (late): $t(106) = -0.638, p = 0.525, d = -0.126, \mathrm{CI} = [-2.720,
1.362]$; reduced versus reduced (late): $t(100) = -0.407, p = 0.685, d =
-0.082, \mathrm{CI} = [-2.477, 1.626]$). These comparisons suggest that recall
accuracy has a directional component: accuracy is affected differently by
removing features later that had been present earlier versus adding features
later that had \textit{not} been present earlier. In contrast, we found that
participants exhibited more temporal and feature-based clustering when we added
incidental visual features to \textit{any} lists (comparisons of clustering on
feature-rich versus reduced lists are reported above; temporal clustering in
reduced versus reduced (early) and reduced versus reduced (late) conditions:
$t$s $\leq -9.885$, $p$s $< 0.001$; feature-based clustering in reduced versus
reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-4.555$, $p$s $< 0.001$). Temporal and feature-based clustering were not
reliably different in the feature-rich, reduced (early), and reduced (late)
conditions (temporal clustering in feature-rich versus reduced (early) and
feature-rich versus reduced (late) conditions: $t$s $\geq -1.379$, $p$s $\geq
0.171$; feature-based clustering in feature-rich versus reduced (early) and
feature-rich versus reduced (late) conditions: $|t|$s $\leq 1.441$, $p$s $\geq
0.153$).

@@

\caption{\textbf{Recall dynamics in feature-rich free recall (order
manipulation conditions).} \textbf{A.} Behavioral plots. \textbf{Left panels.}
The probabilities of initiating recall with each word are plotted as a function
of presentation position. \textbf{Middle panels.} The conditional probabilities
of recalling each word are plotted as a function of the relative position (Lag)
to the words recalled just-prior. \textbf{Right panels.} The overall
probabilities of recalling each word are plotted as a function of presentation
position. \textbf{All panels.} Error ribbons denote bootstrap-estimated 95\%
confidence intervals (calculated across participants). Top panels display the
recall dynamics for early (order manipulation) lists in each condition (color).
Bottom panels display the recall dynamics for late (randomly ordered) lists.
See Figures~\dynamicsRandom~and~\dynamicsAdaptive~for analogous plots for the
random and adaptive conditions. \textbf{B.} Feature distances ($z$-scored
within condition) between the features of successively presented words (see
\textit{Defining feature-based distances}), for each condition's feature of
focus, plotted as a function of presentation position.\textbf{C.} Proportion of
event boundaries (see \textit{Identifying event boundaries}) for each
condition's feature of focus, plotted as a function of presentation position.}

@@

showed an increase in memory performance on early lists (accuracy, relative to
early feature-rich lists; $t(95) = 3.034, p = 0.003, d = 0.667, \mathrm{CI} =
[1.048, 5.113]$). Participants in the color-ordered condition also showed a

@@

early feature-rich lists: $t(96) = 1.850, p = 0.067, d = 0.402, \mathrm{CI} =
[-0.010, 3.712]$; Fig.~\ref{fig:barplots}A). Participants' performances on

@@

$\leq 1.013, p$s $\geq 0.314$). Participants in both of the semantically

@@

(versus early feature-rich lists; category: $t(95) = 8.813, p < 0.001, d =
1.936, \mathrm{CI} = [6.793, 11.751] $; size: $t(95) = 2.630, p = 0.010, d =
0.578, \mathrm{CI} = [0.831, 4.866]$; Fig.~\ref{fig:barplots}B). Participants

@@

-1.547, p = 0.125, d = -0.340, \mathrm{CI} = [-3.693, 0.341]$), whereas

@@

clustering on early lists ($t(95) = 2.858, p = 0.005, d = 0.628, \mathrm{CI} =
[1.031, 4.886]$). Participants in the visually ordered conditions exhibited
more similar performance (accuracy) on early lists, relative to early feature

@@

color-ordered condition: $t(96) = 1.850, p = 0.067, d = 0.402, \mathrm{CI} =
[-0.010, 3.712]$; location: $t(95) = 0.043, p = 0.966, d = 0.010, \mathrm{CI} =
[-1.598, 1.729]$). Participants in the visually ordered conditions also showed
similar temporal clustering on early lists, relative to early feature-rich
lists (color: $t(96) = -1.339, p = 0.184, d = -0.291, \mathrm{CI} = [-3.238,

@@

condition: $t(95) = 1.705, p = 0.092, d = 0.374, \mathrm{CI} = [-0.155,

@@

condition (category: $t(125) = 2.722, p = 0.007, d = 0.484, \mathrm{CI} =
[0.827, 4.932]$; size: $t(125) = 3.866, p < 0.001, d = 0.687, \mathrm{CI} =
[2.020, 5.983]$), but showed no reliable differences in lexicographic (length:
$t(125) = 0.521, p = 0.603, d = 0.093, \mathrm{CI} = [-1.311, 2.333]$; first
letter: $t(125) = -0.842, p = 0.401, d = -0.150, \mathrm{CI} = [-2.825,
1.095]$) or visual (color: $t(125) = -0.650, p = 0.517, d = -0.116, \mathrm{CI}
= [-2.680, 1.249]$; location: $t(125) = -0.251, p = 0.802, d = -0.045,
\mathrm{CI} = [-2.257, 1.524]$) clustering. Similarly, participants in the

@@

rich participants) lexicographic clustering (length: $t(125) = 3.682, p <
0.001, d = 0.655, \mathrm{CI} = [1.890, 5.569]$; first letter: $t(125) = 5.134,
p < 0.001, d = 0.912, \mathrm{CI} = [3.251, 7.258]$) on early lists, but showed
no reliable differences in semantic (category: $t(125) = -1.040, p = 0.301, d =
-0.185, \mathrm{CI} = [-3.095, 1.092]$; size: $t(125) = 0.006, p = 0.995, d =
0.001, \mathrm{CI} = [-1.933, 1.952]$) or visual (color: $t(125) = 0.092, p =
0.927, d = 0.016, \mathrm{CI} = [-1.834, 1.867]$; location: $t(125) = 0.407, p
= 0.685, d = 0.072, \mathrm{CI} = [-1.655, 2.463]$) clustering. And

@@

color: $t(126) = 2.022, p = 0.045, d = 0.358, \mathrm{CI} = [0.056, 3.965]$;
location: $t(126) = 4.390, p < 0.001, d = 0.777, \mathrm{CI} = [2.730, 6.199]

@@

p = 0.991, d = 0.002, \mathrm{CI} = [-1.988, 1.871]$; size: $t(126) = -0.104, p
= 0.917, d = -0.018, \mathrm{CI} = [-2.166, 1.847] $) or lexicographic (length:
$t(126) = 0.592, p = 0.555, d = 0.105, \mathrm{CI} = [-1.361, 2.420]$; first
letter: $t(126) = 0.040, p = 0.968, d = 0.007, \mathrm{CI} = [-1.791, 1.863]$)

@@

broad patterns (Figs.~\ref{fig:recall-dynamics}A, \ref{fig:fingerprints}).

@@

clustering, and so on). Although it is not possible to fully separate feature

@@

\caption{\textbf{Memory ``fingerprints'' (order manipulation conditions).} The
across-participant clustering scores for each feature type ($x$-coordinate) are
displayed for each experimental condition (color), separately for order
manipulation (early, top) and randomly ordered (late, bottom) lists. Error bars
denote bootstrap-estimated 95\% confidence intervals. See
Figures~\fingerprintsRandom~and~\fingerprintsAdaptive~for analogous plots for
the random and adaptive conditions.} \label{fig:fingerprints}

@@

    \caption{\textbf{Memory fingerprint dynamics (order manipulation
    conditions).} \textbf{A.} Each column (and color) reflects an experimental
    condition. In the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for one list. Order
    manipulation (early) lists are denoted by circles and randomly ordered
    (late) lists are denoted by stars. All of the fingerprints (across all
    conditions and lists) are projected into a common space. The bar plots in
    the bottom panels display the Euclidean distances of the per-list memory
    fingerprints to the average fingerprint across all prior lists, for each
    condition. Error bars denote bootstrap-estimated 95\% confidence intervals.
    The dotted vertical lines denote the boundaries between early and late
    lists. \textbf{B.} In this panel, the fingerprints for early (circle) and
    late (star) lists are averaged across lists and participants before
    projecting the fingerprints into a (new) 2D space. See
    Figure~\fingerprintTrajectoryRandom~for analogous plots for the random
    conditions. } \label{fig:fingerprint-trajectories}

>>>>

To obtain a clean preliminary estimate of the consequences on memory of
randomly varying the font colors and locations of presented words (versus
holding the font color fixed at black, and the words' locations fixed
at the center of the screen), we compared participants' performance on the
\textit{feature-rich} and \textit{reduced} experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In the feature-rich
condition, the words' colors and locations varied randomly, and in
the reduced condition, words were always presented in black, at the center of
the display. Aggregating across all lists for each participant, we found no
difference in recall accuracy (i.e., the proportions of words successfully recalled)
for feature-rich versus reduced lists ($t(126) = -0.290,~p = 0.772$,
Cohen's~$d = -0.051$, bootstrap-estimated 95\% confidence interval (CI)~$= [-2.387,~
1.768]$). However, participants in the feature-rich condition clustered their
recalls substantially more along every dimension we examined (temporal
clustering: $t(126) = 10.632,~p < 0.001,~d = 1.882,~\mathrm{CI} = [7.786,~
14.386]$; semantic category clustering: $t(126) = 10.148,~p < 0.001,~d = 1.796,~
\mathrm{CI} = [7.324,~13.778]$; size clustering: $t(126) = 12.033,~p < 0.001,~
d = 2.129,~\mathrm{CI} = [9.030,~15.918]$; word length clustering: $t(126) =
10.720,~p < 0.001,~d = 1.897,~\mathrm{CI} = [7.442,~15.174]$; first letter
clustering: $t(126) = 6.679,~p < 0.001,~d = 1.182,~\mathrm{CI} = [4.490,~
9.611]$; see \textit{Permutation-corrected feature clustering scores} for more
information about how we quantified each participant's clustering tendencies.)
Taken together, these comparisons suggest that adding new features changes how
participants organize their memories of studied words, even when those new
features are independent of the words themselves and
vary randomly across words. We found no evidence that those additional
uninformative features were distracting (in terms of their impact on memory
performance), but they did affect participants' recall dynamics (measured via
their clustering scores).

A core assumption of our approach is that each participant organizes their
memories in a unique way. We defined each participant's \textit{memory
fingerprint} as the set of their permutation-corrected clustering scores across
all dimensions we tracked in our study, including their six feature-based
clustering scores (category, size, length, first letter, color, and location)
and their temporal clustering score. Conceptually, a participant's memory
fingerprint describes their tendency to order, in their recall sequences (and
presumably, organize in memory), the studied words along each dimension. If
these memory fingerprints are truly unique to each participant, then we would
expect that the estimated fingerprints computed for a given participant, on
different lists, should be more similar than the estimated fingerprints
computed for different participants. We reasoned that the feature-rich
condition would provide the best opportunity to test this assumption, since the
clustering scores would not be potentially confounded by order manipulations.
To test our ``unique memory fingerprint'' assumption, we compared the
similarity (correlation) between the fingerprint from a single list (from one
participant) and (a) the average fingerprint from all other lists from the same
participant versus (b) the average fingerprint from each other participant
(across all of their lists). Repeating this procedure for all lists and
participants, we found that participants' fingerprints for a held-out list were
reliably more similar to their fingerprints for other lists than they were to
 other participants' fingerprints ($t(70280) = 5.077,~p < 0.001,~d =
0.162,~\mathrm{CI} = [3.086,~6.895]$). This suggests that participants'
fingerprints are relatively stable across lists, and that each participant's fingerprint
is unique to them.

We next asked whether adding these incidental visual features to later lists
(after the participants had already studied impoverished lists), or removing
the visual features from later lists (after the participants had already
studied visually diverse lists) might affect memory performance. In other
words, we sought to test for potential effects of changing the ``richness'' of
participants' experiences over time. All participants studied and recalled a
total of 16 lists; we defined \textit{early} lists as the first eight lists and
\textit{late} lists as the last eight lists each participant encountered. To
help interpret our results, we compared participants' memories on early versus
late lists in the above feature-rich and reduced conditions. Participants in
both conditions remembered more words on early versus late lists (feature-rich:
$t(66) = 4.553,~p < 0.001,~d = 0.233,~\mathrm{CI} = [2.427,~7.262]$; reduced:
$t(60) = 2.434,~p = 0.018,~d = 0.134,~\mathrm{CI} = [0.493,~4.910]$).
Participants in the feature-rich (but not reduced) conditions exhibited more
temporal clustering on early versus late lists (feature-rich: $t(66) = 2.268,~p
= 0.027,~d = 0.181,~\mathrm{CI} = [0.437,~4.425]$; reduced: $t(60) = 0.986,~p =
0.328,~d = 0.061,~\mathrm{CI} = [-0.897,~3.348]$). And participants in both
conditions tended to exhibit more semantic clustering on early versus late
lists (feature-rich, category: $t(66) = 3.684,~p < 0.001,~d = 0.220,~
\mathrm{CI} = [1.733,~5.732]$; feature-rich, size: $t(66) = 1.629,~p = 0.108,~d
= 0.100,~\mathrm{CI} = [-0.207,~3.905] $; reduced, category: $t(60) = 2.755,~p
= 0.008,~d = 0.177,~\mathrm{CI} = [0.761,~5.189]$; reduced, size: $t(60) =
3.081,~p = 0.003,~d = 0.201,~\mathrm{CI} = [1.210,~5.326]$). Participants in
the reduced (but not feature-rich) conditions tended to exhibit more
lexicographic clustering on early versus late lists (feature-rich, word length:
$t(66) = -0.100,~p = 0.921,~d = -0.010,~\mathrm{CI} = [-2.217,~1.899]$; feature
rich, first letter: $t(66) = 0.412,~p = 0.681,~d = 0.045,~\mathrm{CI} =
[-1.645,~2.461]$; reduced, word length: $t(60) = 3.762,~p < 0.001,~d = 0.261,~
\mathrm{CI} = [1.604,~6.821]$; reduced, first letter: $t(60) = 1.721,~p =
0.090,~d = 0.175,~\mathrm{CI} = [-0.138,~4.098]$). Taken together, these
comparisons suggest that even when the presence or absence of incidental visual
features is stable across lists, participants still exhibit some differences in
their performance and memory organization tendencies for early versus late
lists.

With these differences in mind, we next compared participants' memories on
early versus late lists for two additional experimental conditions (see
\textit{Random conditions}, Fig.~\dynamicsRandom). In a \textit{reduced
(early)} condition, we held the visual features constant on early lists, but
allowed them to vary randomly on late lists. In a \textit{reduced (late)}
condition, we allowed the visual features to vary randomly on early lists, but
held them constant on late lists. Given our above findings that (a)
participants tended to exhibit stronger clustering effects on feature-rich
(versus reduced) lists, and (b) participants tended to remember more words and
exhibit stronger clustering effects on early (versus late) lists, we expected
these early versus late differences to be enhanced in the reduced (early)
condition and diminished in the reduced (late) condition. However, to our
surprise, participants in \textit{neither} condition exhibited reliable early-versus-late
differences in accuracy (reduced (early): $t(41) = 1.499, ~p =
0.141,~d = 0.098,~\mathrm{CI} = [-0.345,~3.579]$; reduced (late): $t(40) =
1.462,~p = 0.152,~d = 0.121,~\mathrm{CI} = [-0.376,~2.993]$), temporal
clustering (reduced (early): $t(41) = 0.857,~p = 0.396,~d = 0.068,~\mathrm{CI}
= [-1.012,~2.896]$; reduced (late): $t(40) = 1.244,~p = 0.221,~d = 0.128,~
\mathrm{CI} = [-0.894,~3.088]$), nor feature-based clustering (reduced (early),
category: $t(41) = 0.707,~p = 0.484,~d = 0.068,~\mathrm{CI} = [-1.314,~2.830]$;
reduced (early), size: $t(41) = 0.803,~p = 0.427,~d = 0.079,~\mathrm{CI} =
[-1.142,~2.953]$; reduced (early), length: $t(41) = 0.461,~p = 0.648,~d =
0.060,~\mathrm{CI} = [-1.545,~2.462]$; reduced (early), first letter: $t(41) =
0.781,~p = 0.439,~d = 0.101,~\mathrm{CI} = [-1.039,~2.881]$; reduced (late),
category: $t(40) = 0.101,~p = 0.920,~d = 0.009,~\mathrm{CI} = [-1.776,~2.307]$;
reduced (late), size: $t(40) = 0.555,~p = 0.582,~d = 0.058,~\mathrm{CI} =
[-1.444,~2.274]$; reduced (late), length: $t(40) = 1.482,~p = 0.146,~d =
0.126,~\mathrm{CI} = [-0.444,~3.743]$; reduced (late), first letter: $t(40) =
-0.143,~p = 0.887,~d = -0.017,~\mathrm{CI} = [-2.204,~1.830] $). We
hypothesized that adding or removing the variability in the visual features was
acting as a sort of ``event boundary'' between early and late
lists~\citep[e.g.,][]{ClewEtal19,RadvZack17,RadvCope06}. In prior work, we
(and others) have found that memories formed just after event boundaries can be
enhanced~\citep[e.g., due to less contextual interference between pre- and
post-boundary items;][]{MannEtal16, PettEtal16, GoldEtal17, FlorEtal17}.

We found that \textit{adding} incidental visual features on later lists that
had not been present on early lists (as in the reduced (early) condition)
served to enhance recall performance relative to conditions where all lists had
the same blends of features (accuracy for feature-rich vs.~reduced (early):
$t(107) = -2.230,~p = 0.028,~d = -0.439,~\mathrm{CI} = [-4.252,~-0.229]$;
reduced vs.~reduced (early): $t(101) = -2.045,~p = 0.043,~d = -0.410,~
\mathrm{CI} = [-3.826,~0.112]$; also see Fig.~\accuracyByList A). However,
\textit{subtracting} irrelevant visual features on later lists that
\textit{had} been present on early lists (as in the reduced (late) condition)
did not appear to impact recall performance (accuracy for feature-rich vs.~
reduced (late): $t(106) = -0.638,~p = 0.525,~d = -0.126,~\mathrm{CI} = [-2.720,~
1.362]$; reduced vs.~reduced (late): $t(100) = -0.407,~p = 0.685,~d =
-0.082,~\mathrm{CI} = [-2.477,~1.626]$). These comparisons suggest that recall
accuracy has a directional component: accuracy is affected differently by
removing features that had initially been present versus adding features
that had initially been absent. In contrast, we found that
participants exhibited more temporal and feature-based clustering when we added
incidental visual features to \textit{any} lists (comparisons of clustering on
feature-rich vs.~reduced lists are reported above; temporal clustering in
reduced vs.~reduced (early) and reduced vs.~reduced (late) conditions:
$t$s $\leq -9.885$,~$p$s $< 0.001$; feature-based clustering in reduced versus
reduced (early) and reduced versus reduced (late) conditions: $t$s $\leq
-4.555$,~$p$s $< 0.001$). Temporal and feature-based clustering were not
reliably different in the feature-rich, reduced (early), and reduced (late)
conditions (temporal clustering in feature-rich vs.~reduced (early) and
feature-rich vs.~reduced (late) conditions: $t$s $\geq -1.379$,~$p$s $\geq
0.171$; feature-based clustering in feature-rich vs.~reduced (early) and
feature-rich vs.~reduced (late) conditions: $|t|$s $\leq 1.441$,~$p$s $\geq
0.153$).

@@

\caption{\textbf{Recall dynamics in feature-rich free recall (order
manipulation conditions).} \textbf{A.} Behavioral plots. \textbf{Left panels.}
The probabilities of initiating recall with each word are plotted as a function
of presentation position. \textbf{Middle panels.} The conditional probabilities
of recalling each word are plotted as a function of the relative position (lag)
to the word recalled just-prior. \textbf{Right panels.} The overall
probabilities of recalling each word are plotted as a function of presentation
position. \textbf{All panels.} Error ribbons denote bootstrap-estimated 95\%
confidence intervals (calculated across participants). Top panels display the
recall dynamics for early (order manipulation) lists in each condition (color).
Bottom panels display the recall dynamics for late (randomly ordered) lists.
See Figures~\dynamicsRandom~and~\dynamicsAdaptive~for analogous plots for the
random and adaptive conditions. \textbf{B.} Distances between successively
presented words ($z$-scored within condition) computed based on each condition's
feature of focus, and plotted as a function of presentation position. See
\textit{Defining feature-based distances} for additional information. \textbf{C.} Proportion of
event boundaries (see \textit{Identifying event boundaries}) for each
condition's feature of focus, plotted as a function of presentation position.}

@@

showed an increase in memory performance on early lists (accuracy relative to
early feature-rich lists; $t(95) = 3.034,~p = 0.003,~d = 0.667,~\mathrm{CI} =
[1.048,~5.113]$). Participants in the color-ordered condition also showed a

@@

early feature-rich lists: $t(96) = 1.850,~p = 0.067,~d = 0.402,~\mathrm{CI} =
[-0.010,~3.712]$; Fig.~\ref{fig:barplots}A). Participants' performances on

@@

$\leq 1.013,~p$s $\geq 0.314$). Participants in both of the semantically

@@

(vs.~early feature-rich lists; category: $t(95) = 8.813,~p < 0.001,~d =
1.936,~\mathrm{CI} = [6.793,~11.751] $; size: $t(95) = 2.630,~p = 0.010,~d =
0.578,~\mathrm{CI} = [0.831,~4.866]$; Fig.~\ref{fig:barplots}B). Participants

@@

-1.547,~p = 0.125,~d = -0.340,~\mathrm{CI} = [-3.693,~0.341]$), whereas

@@

clustering on early lists ($t(95) = 2.858,~p = 0.005,~d = 0.628,~\mathrm{CI} =
[1.031,~4.886]$). Participants in the visually ordered conditions exhibited
more similar performance (accuracy) on early lists relative to early feature

@@

color-ordered condition: $t(96) = 1.850,~p = 0.067,~d = 0.402,~\mathrm{CI} =
[-0.010,~3.712]$; location: $t(95) = 0.043,~p = 0.966,~d = 0.010,~\mathrm{CI} =
[-1.598,~1.729]$). Participants in the visually ordered conditions also showed
similar temporal clustering on early lists relative to early feature-rich
lists (color: $t(96) = -1.339,~p = 0.184,~d = -0.291,~\mathrm{CI} = [-3.238,~

@@

condition: $t(95) = 1.705,~p = 0.092,~d = 0.374,~\mathrm{CI} = [-0.155,~

@@

condition (category: $t(125) = 2.722,~p = 0.007,~d = 0.484,~\mathrm{CI} =
[0.827,~4.932]$; size: $t(125) = 3.866,~p < 0.001,~d = 0.687,~\mathrm{CI} =
[2.020,~5.983]$), but showed no reliable differences in lexicographic (length:
$t(125) = 0.521,~p = 0.603,~d = 0.093,~\mathrm{CI} = [-1.311,~2.333]$; first
letter: $t(125) = -0.842,~p = 0.401,~d = -0.150,~\mathrm{CI} = [-2.825,~
1.095]$) or visual (color: $t(125) = -0.650,~p = 0.517,~d = -0.116,~\mathrm{CI}
= [-2.680,~1.249]$; location: $t(125) = -0.251,~p = 0.802,~d = -0.045,
\mathrm{CI} = [-2.257,~1.524]$) clustering. Similarly, participants in the

@@

rich participants) lexicographic clustering (length: $t(125) = 3.682,~p <
0.001,~d = 0.655,~\mathrm{CI} = [1.890,~5.569]$; first letter: $t(125) = 5.134,
p < 0.001,~d = 0.912,~\mathrm{CI} = [3.251,~7.258]$) on early lists, but showed
no reliable differences in semantic (category: $t(125) = -1.040,~p = 0.301,~d =
-0.185,~\mathrm{CI} = [-3.095,~1.092]$; size: $t(125) = 0.006,~p = 0.995,~d =
0.001,~\mathrm{CI} = [-1.933,~1.952]$) or visual (color: $t(125) = 0.092,~p =
0.927,~d = 0.016,~\mathrm{CI} = [-1.834,~1.867]$; location: $t(125) = 0.407, p
= 0.685,~d = 0.072,~\mathrm{CI} = [-1.655,~2.463]$) clustering. And

@@

color: $t(126) = 2.022,~p = 0.045,~d = 0.358,~\mathrm{CI} = [0.056,~3.965]$;
location: $t(126) = 4.390,~p < 0.001,~d = 0.777,~\mathrm{CI} = [2.730,~6.199]

@@

p = 0.991,~d = 0.002,~\mathrm{CI} = [-1.988,~1.871]$; size: $t(126) = -0.104, p
= 0.917,~d = -0.018,~\mathrm{CI} = [-2.166,~1.847] $) or lexicographic (length:
$t(126) = 0.592,~p = 0.555,~d = 0.105,~\mathrm{CI} = [-1.361,~2.420]$; first
letter: $t(126) = 0.040,~p = 0.968,~d = 0.007,~\mathrm{CI} = [-1.791,~1.863]$)

@@

broad patterns (Figs.~\ref{fig:recall-dynamics}A,~\ref{fig:fingerprints}).

@@

clustering, and so on). Although it is not possible to fully separate feature-based

@@

\caption{\textbf{Memory ``fingerprints'' (order manipulation conditions).} The
across-participant average clustering scores for each feature type ($x$-axis) are
displayed for each experimental condition (color), separately for order-manipulated
(early, top) and randomly ordered (late, bottom) lists. Error bars
denote bootstrap-estimated 95\% confidence intervals. See
Figures~\fingerprintsRandom~and~\fingerprintsAdaptive~for analogous plots for
the random and adaptive conditions.} \label{fig:fingerprints}

@@

    \caption{\textbf{Memory fingerprint dynamics (order manipulation
    conditions).} \textbf{A.} Each column (and color) reflects an experimental
    condition. In the top panels, each marker displays a 2D projection of the
    (across-participant) average memory fingerprint for a single list. Order
    manipulation (early) lists are denoted by circles and randomly ordered
    (late) lists are denoted by stars. All of the fingerprints (across all
    conditions and lists) are projected into a common space. The bar plots in
    the bottom panels display the Euclidean distances between each per-list memory
    fingerprint and the average fingerprint across all prior lists, for each
    condition. Error bars denote bootstrap-estimated 95\% confidence intervals.
    The dotted vertical lines denote the boundaries between early and late
    lists. \textbf{B.} In this panel, the fingerprints for early (circle) and
    late (star) lists are averaged across lists and participants before
    projecting the fingerprints into a (new) 2D space. See
    Figure~\fingerprintTrajectoryRandom~for analogous plots for the random
    conditions. } \label{fig:fingerprint-trajectories}
